 

### 2021-07-27 vertical-thermal-review.Rmd
updated

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]. Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].As a result of increasing C allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizerelatedScalingTree2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**? REFS?**).

@@

In contrast, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments and sometimes forest census data.
While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature- sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or high temperatures among large/ canopy trees [**Fig. `r fig_treerings`**, @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021] , corroborating evidence from forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
Despite the potential for shorter trees in open forests to experience greater environmental stress [*Fig. `r fig_schematic`b*; **REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].

>>>>

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical]. Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].As a result of increasing C allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Because $g_s$ tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on time scale of hours, days, or seasons [@christoffersenLinkingHydraulicTraits2016; Garcia et al. 2021],consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2). More active stomatal regulation of tall canopy trees (e.g., Mediavilla & Escudero 2004)-- often combined with greater effective rooting depth--may offset greater xylem embolism risk (Olson et al. 2018, Garcia et al. 2021, Chitra-Tarak et al. 2021) and lower capacity to adjust traits related to carbon metabolism to drought (Bartholomew et al., 2020). This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height (Table 1; McDowell 2011).

@@

The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature- sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or high temperatures among large/ canopy trees [**Fig. `r fig_treerings`**, @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021] , corroborating evidence from forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
Despite the potential for shorter trees in open forests to experience greater environmental stress [*Fig. `r fig_schematic`b*; **REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
